\section{Object classifiers}

The most basic classifying types are univalent universes. Recall that the universe comes equipped with a \emph{universal type family}. This type family is so-called because any family of small types can be given as a map into the universe. Moreover, this representation is essentially unique by the univalence axiom. Indeed, if we are given a family of equivalences
\begin{equation*}
  \prd{x:A}B(x)\simeq C(x)
\end{equation*}
between two type families over a type $A$, then for any univalent universe $\UU$ containing both families $B$ and $C$ it follows that the maps $B,C:A\to\UU$ are homotopic, and therefore can be identified. To arrive at the definition of an object classifier, we phrase the above observations in terms of pullbacks.

\subsection{Univalent universes are object classifiers}

\begin{defn}
  Consider a map $p:E\to B$ and a map $f:X\to Y$. The type $\cart(f,p)$ of \define{cartesian morphisms} from $f$ to $p$ is the type of quadruples $(g,h,H,t)$ consisting of maps
  \begin{align*}
    g & : Y\to B \\
    h & : X\to E,
  \end{align*}
  a homotopy $H:g\circ f\htpy p\circ h$, and a term $t$ witnessing that the commuting square
  \begin{equation*}
    \begin{tikzcd}
      X \arrow[r,"h"] \arrow[d,swap,"f"] & E \arrow[d,"p"] \\
      Y \arrow[r,swap,"g"] & B
    \end{tikzcd}
  \end{equation*}
  is a pullback square.
\end{defn}

\begin{defn}
  A map $p:E\to B$ is called an \define{object classifier} if for every map $f:X\to Y$, the type $\cart(f,p)$ of cartesian morphisms
  \begin{equation*}
    \begin{tikzcd}
      X \arrow[r] \arrow[d,swap,"f"] & E \arrow[d,"p"] \\
      Y \arrow[r] & B
    \end{tikzcd}
  \end{equation*}
  from $f$ to $p$ is a proposition. 
\end{defn}

\begin{prp}
  Consider a map $p:E\to B$. The following are equivalent:
  \begin{enumerate}
  \item The map $p$ is an object classifier.
  \item The function
    \begin{equation*}
      \tr_{\fibf{p}}:(x=y)\to (\fib{p}{x}\simeq\fib{p}{y})
    \end{equation*}
    is an equivalence.
  \end{enumerate}
\end{prp}

\begin{cor}
  A universe is an object classifier if and only if it is univalent.
\end{cor}
