\section{The wedge and smash product of pointed types}

\begin{defn}
Let $A$ and $B$ be pointed types.
\begin{enumerate}
\item We define the \define{wedge} $A\vee B$ of $A$ and $B$ to be the pushout
\begin{equation*}
\begin{tikzcd}
\unit \arrow[r] \arrow[d] & B \arrow[d] \\
A \arrow[r] & A\vee B
\end{tikzcd}
\end{equation*}
\item We define the \define{smash product} $A\wedge B$ of $A$ and $B$ to be the cofiber of the cogap map of the square
\begin{equation*}
\begin{tikzcd}
\unit \arrow[r] \arrow[d] & B \arrow[d] \\
A \arrow[r] & A\times B
\end{tikzcd}
\end{equation*}
That is, the smash product is defined as the cofiber of the canonical map $A\vee B\to A\times B$. 
\end{enumerate}
\end{defn}

For any two pointed types $A$ and $B$, there is a pointed map
\begin{equation*}
\mathsf{pair}_\ast : A \to_\ast (B\to_\ast A\wedge B).
\end{equation*}

\begin{thm}\label{thm:smash_adj}
Let $A$, $B$, and $X$ be pointed types. Then the pointed map
\begin{equation*}
(A\wedge B \to_\ast X)\to_\ast (A \to_\ast (B\to_\ast X))
\end{equation*}
given by $f\mapsto f\mathbin{\circ_\ast}\mathsf{pair}_\ast$ is an equivalence.
Moreover, these equivalences are natural in $A$, $B$, and $X$ in the sense that...
\end{thm}

\begin{cor}
For any $m,n:\N$ we have an equivalence
\begin{equation*}
\eqv{{\sphere{m}}\wedge{\sphere{n}}}{\sphere{m+n}}.
\end{equation*}
\end{cor}

\begin{proof}
We have
\begin{align*}
({\sphere{m}}\wedge{\sphere{n}}\to_\ast X) & \eqvsym (\sphere{m}\to_\ast (\sphere{n}\to_\ast X)) \\
& \eqvsym \loopspace[m]{\loopspace[n]{X}} \\
& \eqvsym \loopspace[m+n]{X} \\
& \eqvsym (\sphere{m+n}\to_\ast X)
\end{align*}
By the naturality of the equivalences in \cref{thm:smash_adj} it follows that the composite equivalence is given by precomposition by the pointed map 
\begin{equation*}
\sphere{m}\wedge\sphere{n}\to_\ast \sphere{m+n}
\end{equation*}
that corresponds to the identity map $\sphere{m+n}\to_\ast \sphere{m+n}$. Thus it follows by \cref{ex:yoneda_ptd_types} that this pointed map is an equivalence.
\end{proof}

\begin{thm}
Given two pointed spaces, there is an equivalence
\begin{equation*}
\eqv{\join{X}{Y}}{\susp(X\wedge Y)}.
\end{equation*}
\end{thm}

\begin{exercises}
\exercise 
\begin{subexenum}
\item Show that $Y$ is equivalent to the mapping cone of $X\to X\vee Y$.
\item Show that the pushout of $X \leftarrow X\vee Y \rightarrow Y$ is contractible.
\end{subexenum}
\exercise Let $A$ and $B$ be pointed types. Show that the square
\begin{equation*}
\begin{tikzcd}
A+B \arrow[r] \arrow[d] & A\times B \arrow[d] \\
1+1 \arrow[r] & A\wedge B
\end{tikzcd}
\end{equation*}
is cocartesian.
\exercise Show that if
\begin{equation*}
\begin{tikzcd}
S_1 \arrow[r] \arrow[d] & Y_1 \arrow[d] & S_2 \arrow[r] \arrow[d] & Y_2 \arrow[d] \\
X_1 \arrow[r] & Z_1 & X_2 \arrow[r] & Z_2
\end{tikzcd}
\end{equation*}
are pushout squares, where all types, maps and homotopies are pointed, then so is
\begin{equation*}
\begin{tikzcd}
S_1\vee S_2 \arrow[r] \arrow[d] & Y_1\vee Y_2 \arrow[d] \\
X_1 \vee X_2 \arrow[r] & Z_1\vee Z_2. 
\end{tikzcd}
\end{equation*}
\exercise Show that if
\begin{equation*}
\begin{tikzcd}
S \arrow[r] \arrow[d] & Y \arrow[d] \\
X \arrow[r] & Z
\end{tikzcd}
\end{equation*}
is a cocartesian square of pointed spaces, then the cofiber of $X\vee Y\to Z$ is equivalent to $\susp(S)$.
\exercise Show that there is an equivalence
\begin{equation*}
\eqv{\susp(X\times Y)}{\susp(X\vee Y)\vee \susp(X\wedge Y)}
\end{equation*}
\exercise Show that $\susp(X\vee Y)$ is a retract of $\susp(X\times Y)$. 
\exercise Show that if $f:A\to X$ is a constant of pointed spaces, then $\eqv{M_f}{X\vee \susp(A)}$. 
\exercise Show that the cofiber of the diagonal $\delta:\sphere{1}\to \sphere{1}\times\sphere{1}$ is equivalent to $\sphere{2}\vee\sphere{2}$.
\exercise Show that $\eqv{\mathsf{Fin}(n+1)\wedge \mathsf{Fin}(m+1)}{\mathsf{Fin}(n\cdot m)+\unit}$.
\end{exercises}
